Parsing Postfix log files

### 2009-04-11 doc/logparser-results.tex
Corrections to the first coverage section.

@@

The discussion of the parser's coverage of Postfix log files is separated
into two parts: log lines covered and mails covered.  The first is
important because the parser should handle all (relevant) log lines it is
given; the second is equally important because the parser must properly
deal with every mail if it is to be useful.  Improving the former is
less intrusive, as it just requires new rules to be written; improving the
latter is much more intrusive as it requires changes to the parser
algorithm, and it can also be much harder to notice a deficiency.

\subsection{Log Lines Covered}

@@

    \item Check if there are any rules for the Postfix component that
        produced the log line; if not then skip the log line.

    \item Try each rule until a recognising rule is found; if the log line
        is not recognised, issue a warning and move to the next log line.

    \item Execute the action specified by the rule.

@@

Full coverage of log lines requires the following:

    \item Each Postfix component whose log lines are of interest must have
        at least one rule or its log lines will be silently skipped; in the
        extreme case of zero rules the parser would happily skip every log
        line.  There may be any number of log lines from other programs
        intermingled in the log file, and there are some Postfix programs
        that do not produce any log lines of interest.

    \item There must be a rule to recognise each different log line
        produced by each program; if a log line is not successfully
        recognised the parser will issue a warning.  Rules should be as
        specific and tightly bound as possible to ensure accurate
        parsing:\footnote{A rule with a regex that matches zero or more of
        any character will successfully recognise every log line, but not
        in a meaningful way.} most log lines contain fixed strings and have
        a rigid pattern, so this is not a problem.

    \item The appropriate action to take --- discussed in
        \sectionref{mails-covered}.

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser is tested with \numberOFrules{} rules, fully parsing
\numberOFlogFILESall{} contiguous log files from Postfix 2.2 through to
Postfix 2.5), and new rules are easy to add.  Maintaining full coverage is
hard because other servers have different restrictions with custom
messages, \acronym{DNSBL} messages change over time, major releases of
Postfix change warning messages (usually adding more information), etc.,\
so over time the log lines drift and change.  \Graphref{rule hits graph}
shows the number of hits for each rule over all \numberOFlogFILES{} log
files.  The number of hits per rule is quite unevenly spread, resembling a
Power Law distribution --- it is obvious that a small number of rules
recognise the vast majority of the log lines, and more than half the rules
recognise fewer than 100 log lines.

Warnings are issued for any log lines that are not parsed; no warnings are
issued for unparsed log lines while testing with the \numberOFlogFILES{}
test log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which regex parsed it, there
is no way to be sure that every line was parsed by the correct regex,
making it impossible to quantify the false positive rate; however a random
sample of 6,039 log lines was parsed and the results checked manually to
ensure that the correct regex parsed each log line.\footnote{Each log
line was examined and the correct regex identified from the
\numberOFrules{} rules in the database; the correct regex was then
compared to the regex that was used by the parser.}  The sample was
generated by running the following command:

\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!

\noindent{}to randomly extract roughly one line in every 10,000.  Although
on initial appearances exercising only 36 rules (from a total of
\numberOFrules{}) when parsing 6039 log lines seems quite low, after
examining \graphref{rule hits graph} it becomes apparent that such
a low hit rate is to be expected; the reader should also bear in mind that
even when parsing all \numberOFlogFILES{} log files not all the rules are
exercised (some of the rules are for parsing log lines that only appear in
other log files).

>>>>

XXX IS THERE A BETTER WORD THAN ``coverage''?

@@

The discussion of \parsernames{} coverage of Postfix log files is separated
into two parts: log lines correctly recognised, and mail delivery attempts
correctly understood --- the former is a requirement for the latter to be
achieved.  Correctly understanding and reconstructing every mail delivery
attempt, whether it was successful or not, is important so that the
information in the database is correct and complete.  Improving the
proportion of log lines correctly recognised is the less difficult of the
two, because usually it just requires new rules to be written or existing
rules to be changed.  Improving the proportion of correctly understood and
reconstructed mail delivery attempts is more difficult and intrusive,
because it requires adding or changing actions, and it can be much harder
to realise that a deficiency exists and needs to be addressed.

\subsection{Log Lines Correctly Recognised}

@@

    \squeezeitems{}

    \item Skip the log line if the ruleset does not contain any rules for
        the Postfix component that produced the log line.

    \item Try each rule until a recognising rule is found; if the log line
        is not recognised, issue a warning and move on to the next log
        line.

    \item Invoke the action specified by the recognising rule.

@@

Correctly recognising all log lines requires that each Postfix component
whose log lines are of interest must have at least one rule, or its log
lines will be silently skipped; in the extreme case of an empty ruleset the
parser would skip every log line.  \parsername{} skips those log lines
because there may be any number of log lines from other programs
intermingled in the log file, and there are some Postfix components that do
not produce any log lines of interest.  There must be a rule to recognise
each log line variant produced by each Postfix component; if a log line is
not recognised the parser will issue a warning, to inform the user that
they need to extend their ruleset.  Currently, \parsername{} does not parse
log lines from non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it
could easily be extended to do so, if a method could be developed to
correctly associate such log lines with existing state table entries.  Each
rule's regex should be as specific and precise as possible, to ensure
accurate parsing: a rule with a regex that matches zero or more of any
character will recognise every log line, but not in a meaningful way.  Most
log lines contain fixed strings, so this is not a problem in practice.

Full coverage of log lines is easy to achieve, yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files; \parsername{}
has \numberOFrules{} rules, fully parsing \numberOFlogFILESall{} log files
from Postfix 2.2 through to Postfix 2.5.  Maintaining full coverage is hard
because log lines change over time, e.g.\ administrators add restrictions
with custom messages, \acronym{DNSBL} messages change, or major releases of
Postfix change log lines (usually adding more information).  Warnings are
issued for any log lines that are not recognised; no warnings are issued
for unrecognised log lines while parsing the \numberOFlogFILES{} log files,
so it can be safely concluded that there are zero false negatives.  False
positives are harder to quantify: short of examining each of the 60,721,709
log lines and determining which rule recognised it, there is no way to be
sure that every log line was recognised by the correct rule, making it
impossible to quantify the false positive rate.  However, a random sample
of 6,039 log lines was parsed, and the results manually verified by
inspection to ensure that the correct rule recognised each log line.  The
sample was generated by running the command: \texttt{perl -n -e
\singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
randomly extract roughly one log line in every 10,000 (it actually
extracted 0.00994\% instead of 0.001\%).  Each log line was examined and
the correct rule identified from the \numberOFrules{} rules in the
database; the correct rule was then compared to the rule that recognised
the log line when parsing.  The sample results contained zero false
positives, and this check has been automated to ensure continued accuracy.
Based on this, the author is confident that zero false positives occur when
parsing the \numberOFlogFILES{} log files.  On initial appearances,
exercising only 36 rules from a total of \numberOFrules{} when parsing
6,039 log lines seems low, but after examining \graphref{rule hits graph}
it becomes apparent that parsing using such a low number of rules is to be
expected.  The reader should also bear in mind that even when parsing all
\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
because some of the rules are for recognising log lines that only appear in
other log files.
